# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
``` {r echo = TRUE}
activityData = read.csv("activity.csv", colClasses = c("numeric", "Date", "numeric"))
head(activityData)
```
## What is mean total number of steps taken per day?
Ignore the missing values in the dataset.
``` {r echo = TRUE}
completeActivityData = activityData[complete.cases(activityData),]
```
1 - Histogram of the total number of steps taken each day
  - Calculating the total steps per day
``` {r echo = TRUE}
TotalStepsPerDay = aggregate(steps~date,data=completeActivityData,sum)
head(TotalStepsPerDay)
```
Plotting the Histogram showing the Total number of steps taken each day
```{r echo = TRUE}
barplot(TotalStepsPerDay$steps,
     xlab="Day",ylab="Total number of steps taken each day",
     main="Histogram of Total number of steps taken each day",
     names.arg=TotalStepsPerDay$date
     )
```
2 - The mean and median total number of steps taken per day

- The mean of the total number of steps taken per day is 
```{r echo = TRUE} 
mean(TotalStepsPerDay$steps)
```
- The median of the total number of steps taken per day is 
```{r echo = TRUE} 
median(TotalStepsPerDay$steps)
```

## What is the average daily activity pattern?
Calculating the average steps taken accross all days 
```{r echo = TRUE}
avgStepsPerDay = aggregate(steps~interval,data=completeActivityData,mean)
```
1 - Ploting the time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r echo = TRUE}
plot(x = avgStepsPerDay$interval, y = avgStepsPerDay$steps, 
     main="Average number of steps taken averaged accross all days",
     type = "l", col = "grey",
     xlab = "Interval", ylab = "Average steps")

```     

2 - 5-minute interval containing the maximum number of steps
``` {r echo = TRUE}
avgStepsPerDay$interval[which.max(avgStepsPerDay$steps)]
```

## Imputing missing values
1 - The total number of missing values in the dataset 
``` {r echo = TRUE}
length(which(is.na(activityData$steps)))
```

2 - The strategy is to fill in all of the missing values in the dataset with mean value of steps

3 - Creating a new dataset that is equal to the original dataset but with the missing data filled in.
``` {r echo = TRUE, cache=TRUE}
filledActivityData = as.data.frame(activityData)
for (i in 1 : length(filledActivityData$interval)){
        if (is.na(filledActivityData$steps[i])){
                index = which(avgStepsPerDay$interval == filledActivityData$interval[i])
                meanVal = avgStepsPerDay$steps[index]
                filledActivityData$steps[i] = meanVal
        }
}
```

4 - Histogram of the total number of steps taken each day with filled data
```{r echo = TRUE}
TotalStepsPerDay_filled = aggregate(steps~date,data=filledActivityData,sum)
barplot(TotalStepsPerDay_filled$steps,
     xlab="Day",ylab="Total number of steps taken each day",
     main="Histogram of Total number of steps taken each day with filled data",
     names.arg=TotalStepsPerDay_filled$date
     )
```

- The mean of the total number of steps taken per day using filled data is 
```{r echo = TRUE} 
mean(TotalStepsPerDay_filled$steps)
```
- The median of the total number of steps taken per day using filled data is 
```{r echo = TRUE} 
median(TotalStepsPerDay_filled$steps)
```

## Are there differences in activity patterns between weekdays and weekends?
* Create a new factor variable in the dataset with two levels – “weekday” and “weekend” 
``` {r echo = TRUE, cache=TRUE}
weekday = weekdays(filledActivityData$date)

for (i in 1 : length(weekday)){
        if (weekday[i] == "Sunday" | weekday[i] == "Saturday"){
                weekday[i] = "weekend"
                } 
        else  {
                weekday[i] = "weekday" 
                }
}
filledActivityData$day = weekday
```


* Calculating the average steps taken accross weekdays and weekends 
```{r echo = TRUE}
# subsetting data to weekday and weekend
weekdayData = subset(filledActivityData, filledActivityData$day == "weekday")
weekendData = subset(filledActivityData, filledActivityData$day == "weekend")


avgStepsPerDayWeekend = aggregate(steps~interval,data=weekendData,mean)
avgStepsPerDayWeekday = aggregate(steps~interval,data=weekdayData,mean)


```

* Plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken
```{r echo = TRUE}

par(mfrow = c(2, 1))
plot(x = avgStepsPerDayWeekend$interval, y = avgStepsPerDayWeekend$steps, 
    type = "l", xlab = "Interval", ylab = "Average number of steps", main = "Weekend")

plot(x = avgStepsPerDayWeekday$interval, y = avgStepsPerDayWeekday$steps, 
    type = "l", xlab = "Interval", ylab = "Average number of steps", main = "Weekday")
```